---
title: "R Notebook - Jerrie Feng"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

Task 5 in R basics: https://bcb420-2022.github.io/R_basics/r-install.html#task-4---create-your-first-notebook-using-docker

I had to add `chunk_output_type: console` above because the plot would keep showing up inline instead on in the plot window (because of html_notebook option). This caused an error of `plot.new has not been called yet`. Therefore, I had to make it plot in the windows to fix the error.  

```{r}

library() #shows all packages

search() #see which ones are loaded

?vignette #? is to get help

#install seqinr
install.packages("seqinr",repos="https://cran.rstudio.com/")  

library(seqinr) #load package seqinr
ls("package:seqinr") #all functions
?seqinr::a 
seqinr::a("Tyr") #converts AA codon into 1 letter

?seqinr::words #get all possible combo of words from given alphabet
seqinr::words(3, c("A", "G", "C", "U"))

```

Results for `seqinr::words(3, c("A", "G", "C", "U"))` is > seqinr::words(3, c("A", "G", "C", "U"))
 [1] "AAA" "AAG" "AAC" "AAU" "AGA" "AGG" "AGC" "AGU"
 [9] "ACA" "ACG" "ACC" "ACU" "AUA" "AUG" "AUC" "AUU"
[17] "GAA" "GAG" "GAC" "GAU" "GGA" "GGG" "GGC" "GGU"
[25] "GCA" "GCG" "GCC" "GCU" "GUA" "GUG" "GUC" "GUU"
[33] "CAA" "CAG" "CAC" "CAU" "CGA" "CGG" "CGC" "CGU"
[41] "CCA" "CCG" "CCC" "CCU" "CUA" "CUG" "CUC" "CUU"
[49] "UAA" "UAG" "UAC" "UAU" "UGA" "UGG" "UGC" "UGU"
[57] "UCA" "UCG" "UCC" "UCU" "UUA" "UUG" "UUC" "UUU"

```{r}

#get avalible data & load it
data(package="seqinr")
data(aaindex, package="seqinr")

#we get 2 indices from the data
aaindex$FASG890101 
aaindex$PONJ960101

#and then plot it & add labels/text to the plot
#function 'text' did not work at first, had to modify
#showed error of 'a' not being a function
plot(aaindex$FASG890101$I,
     aaindex$PONJ960101$I,
     xlab="hydrophobicity", ylab="volume", type="n")
text(aaindex$FASG890101$I,
     aaindex$PONJ960101$I,
     labels=seqinr::a(names(aaindex$FASG890101$I)))

#to check if the package is installed (only install if it's not there)
if (! requireNamespace("seqinr", quietly=TRUE)) {
  install.packages("seqinr")
}




```



